\section{Introduction}

\Idris{} is a functional programming language with full dependent
types. It is similar to \Epigram~\cite{view-left} or
\Agda{}~\cite{ulf-thesis}, in that it supports dependent pattern
matching. It is built on top of the \Ivor{}~\cite{ivor} theorem
proving library, and is a pure functional language with a syntax
similar to Haskell with GADTs.  The purpose of the language is to
provide a platform for practical programming with dependent types.

\subsection{Motivation}

Several tools exist today for programming with dependent types, in
various forms. These range from extensions to existing languages, such
as Haskell's GADTs~\cite{gadts} and open type
functions~\cite{opentype-haskell} to languages designed with dependent
types from the ground up, such as Cayenne~\cite{cayenne-icfp} and
\Epigram{}~\cite{view-left}. Additionally, theorem provers such as
\Agda{}~\cite{ulf-thesis} and \Coq{}~\cite{coq-manual} are based on
dependent types.

None of these systems, however, are suitable for writing realistic
programs with full dependent types. The type system is either limited
in some way, restricting the type level functions one can
write and thus the properties which can be
shown~\cite{lightweight-dtp}, or the focus is on total correctness,
which, while clearly desirable, can make many simpler and less
safety-critical programs more difficult to write than necessary.

In contrast, we are interested in the \remph{practical} aspects of
dependently typed programming --- we require input and output,
concurrency, and foreign function calls. We also acknowledge that at
times, particularly in the early stages of software development, it
may be necessary to relax some restrictions on the completeness of
proofs. Application development and testing should not stall because
of a missing lemma!  Using our own implementation rather than an
existing tool gives complete freedom to experiment with these
abstractions and language features beyond the type system, such as I/O
and concurrency. Additionally, although unrelated to the work we
present in this paper, the core language is intended as an important
step towards a fully dependently typed implementation of
Hume~\cite{Hume-GPCE}.

\subsection{Contributions}

This paper introduces \Idris{}, a language with dependent types, which
compiles to executable code via C. The paper makes the following
contributions:

\begin{enumerate}
\item We give an implementation of an I/O monad using dependent types,
  in the style of Hancock and Setzer~\cite{hancock-interactive}, which
  compiles to C, and allows foreign functions (implemented in C) to be
  executed without requiring any extensions to the language or type
  system (Section \ref{ioforeign}).
\item We exploit the fact that \Idris{} is built on top of a tactic
  based theorem prover, \Ivor{}~\cite{ivor}, to show how programming
  considerations and proof obligations can be neatly separated
  (Section \ref{interactive}).
\end{enumerate}

Additionally, we demonstrate through a larger example, a verified
implementation of binary arithmetic which we have previously
implemented~\cite{tfp-ccc} in \Ivor{} that a practical, realistic and
efficient dependently typed programming language can be built with
existing technology (Section \ref{sec:binary}).

